### 2012-01-26 contents.tex
Revision after meeting with Christina. Citing 4-eyes report as motivation and future work.

Bug repositories (or bug tracking systems) have since a long time been used in software projects to support coordination among stakeholders. Such systems record discussion and progress of software evolution activities, such as corrective, perfective, and ... changes \cite{Antoniol2008}. Hence, bug repositories are an opportunity to researchers who intend to investigate issues related to the quality of the product and of the process of a software development team, which can give insights about the nature of software development itself.

The information contained in bug repositories have been used in order to predict fault-proneness, to unveil developers' roles from data, to characterize transfer of work in software projects, and so on. Also, it has been used to investigate beliefs in software engineering related to software quality, such as the impact of developer turnover and of collective ownership on the occurrence of bugs.

@@

\TODO why the VERIFIED status is so important?

In summary, for researchers who intend to use data from bug tracking systems, it is not sufficient to understand the data fields in a bug report the way they are documented in the system's documentation. The research should also investigate whether the data fields are used often, whether they are used accordingly to the documentation, and even if the data fields are used consistently across the developers.

In this paper, we investigate the status of a bug, a data field present in most bug tracking systems that reports the progress of the bug solving activity. In particular, we focus on the activity of verification of a fix. We are interested in answering the following exploratory research questions:

\begin{itemize}
	\item How is the verification actually performed? Are there performed ad hoc tests, automated tests, code inspection?
	\item When is the verification performed? Is it performed just after the fix, or is there a specific period for the verification of fixed bugs?
	\item Who performs the verification? Is there a dedicated team for QA? 
	% \item How is the distribution of verification across developers?
	% \item Is there a dedicated team for QA?
\end{itemize}

For research question xx, qualitative of a small sample. 
For research question xx, all bugs were considered.

@@

Bug tracking systems allow users and developers of a software system to manage a list of bugs for the system. Usually, users and developers can report bugs, along with information such as steps to reproduce the bug, its severity and the operating system used. Developers choose bugs to fix and can report on the progress of the bug fixing activities, ask for clarification, discuss causes for the bug etc.

In this research, we focus on Bugzilla, an open source bug tracking system used by notable software projects such as Eclipse, Mozilla, Linux Kernel, Open Office, Apache, and companies such as NASA and Facebook\footnote{Complete list available at http://www.bugzilla.org/installation-list/.}. However, the general concepts from Bugzilla should apply to most other bug tracking systems.

@@

In the simplest cases, a bug is created and receive the status \texttt{UNCONFIRMED} (if it was created by a regular user) or \texttt{NEW} (if it was created by a developer). Next, it is ASSIGNED to a developer, and then it is RESOLVED, possibly by fixing it with a patch on the source code. The solution is then VERIFIED by someone in the quality assurance team. If it passes the quality requirements, it is CLOSED when the next release of the software is released. If it does not pass the quality requirements or if the solution only partially fixes the problem, the bug is REOPENED.

@@

In order to understand the bug fixing process of a software team, however, it should be possible to know what activities are performed upon change of status. For example, what steps are made before changing the status of a bug to RESOLVED (fixed)? Has the patch been applied to the version control system or is it presented as diff file attached to the bug report? Has the patched source code been built and made available for end users (e.g., as a nightly build)? The documentation for Bugzilla does not prescribe specific activities for marking a bug as RESOLVED, relegating instead this decision to each software project.

Similarly, what do developers do before marking a bug as VERIFIED? Bugzilla documentation states that, when a bug is VERIFIED, it means that ``QA [quality assurance team] has looked at the bug and the resolution and agrees that the appropriate resolution has been taken''\footnote{\url{https://landfill.bugzilla.org/bugzilla-3.6-branch/page.cgi?id=fields.html}}. Again, the definition is loose, although it assumes the existence of a QA role.

@@

While mining bug repositories, one cannot assume that the VERIFIED status comprises all forms of software verification. Also, s/he cannot rely on the information provided by the bug tracking system or the developer documentation for a project, since it can be unspecific. Such information can only be assessed by taking a close look on actual bug reports.

@@

	In order to perform the desired analysis, we needed access to all data recorded by Bugzilla for a specific project. We have found such data for two projects --- Eclipse and NetBeans --- from the domain of integrated development environments. The data were made available as part of the Mining Software Repositories 2011 Challenge\footnote{\url{http://www.msrconf.org/msr-challenge.html}} in the form of MySQL database dumps. The files contain all tables from the respective databases, except for the tables containing information about developers.

@@

	NetBeans\footnote{\url{http://netbeans.org/about/history.html}} started  as a student project in 1996, was bought by Sun Microsystems in Oct, 1999, and was open sourced in Mar, 2000.

@@

	There 316,911 bug reports for 155 subprojects in Eclipse, and 185,578 bug reports for 39 subprojects in NetBeans.

@@

\begin{subsection}{Analysis}

	Each of the sampled bugs was read.

@@

\begin{section}{Results and Discussion}
	Some insights were gained in this step:

	\begin{itemize}
		\item In some projects, the status VERIFIED does not correspond to software verification. In Eclipse/EMF, for example, it means that the fix is available in a build that is promoted on the project website\footnote{\url{http://wiki.eclipse.org/Modeling_PMC_Meeting,_2007-10-16}}.
		\item In the NetBeans subprojects that were analyzed, it was common that the reporter also verified the fix. In some bugs, this was necessary since the developer was not able to reproduce the bug.
		\item Often, the fixer verifies the bug just after marking it as fixed.
		\item Sometimes, several old bugs are marked as VERIFIED at the same time.
		\item In Eclipse/Platform, it is common that the fixer requests someone else to verify the fix (``Please verify, Tom'').
		\item In Eclipse/EMF, bug #249436, one developer rewrites the patch submitted by other developer.
		\item In Eclipse/EMF, bug #269789, a test case is used to assess the proposed fix.
	\end{itemize}

@@

	Clearly, in NetBeans the ``verifier'' role is more pronounced. In the subprojects analyzed, approximately 20\% of the developers focus on the verification activity, and they perform about 90\% of the verifications.

	In the Eclipse subprojects, however, such role is not so clear. In EMF, there are no ``verifiers''. That does not mean much because, as we discovered earlier, the VERIFIED status does not correspond to the verification activity. In the case of Eclipse/Platform, only 2\% of the developers are verifiers, and they contribute only to 1\% of the verifications.

@@

\begin{section}{Conclusion}

>>>>

Bug repositories (or bug tracking systems) have for a long time been used in software projects to support coordination among stakeholders. Such systems record discussion and progress of software evolution activities, such as corrective, perfective, and ... changes \cite{Antoniol2008}. Hence, bug repositories are an opportunity to researchers who intend to investigate issues related to the quality of the product and of the process of software development team.
%, or even issues regarding the nature of software development itself.

The information contained in bug repositories has been used in order to predict fault-proneness, to unveil developers' roles from data, to characterize transfer of work in software projects, and so on. Also, it has been used to investigate beliefs in software engineering related to software quality, such as the impact of developer turnover and of collective ownership on the occurrence of bugs.

@@

  In previous research [cite report], we tried to assess the impact of independent verification of bug fixes on software quality, by mining data from bug repositories. 
  We relied on reported verifications tasks, as recorded in bug reports, and interpreted the recorded data according to the documentation for the specific bug tracking system used.
  As the partial results suggested that verification has no impact on software quality, we questioned the accuracy of the data about verification of bug fixes, ad thus decided to investigate how verification is actually performed and reported on specific software projects.

	In particular, we are interested in answer the following exploratory research questions for a limited set of projects:

	\begin{itemize}
		\item \textbf{How} is the verification performed: are there performed ad hoc tests, automated tests, code inspection?
		\item \textbf{Who} performs the verification: is there a dedicated team for QA? 
		\item \textbf{When} is the verification performed: is it performed just after the fix, or is there a verification phase?
		% \item How is the distribution of verification across developers?
		% \item Is there a dedicated team for QA?
	\end{itemize}

% \TODO why the VERIFIED status is so important?

% In summary, for researchers who intend to use data from bug tracking systems, it is not sufficient to understand the data fields in a bug report the way they are documented in the system's documentation. The research should also investigate whether the data fields are used often, whether they are used accordingly to the documentation, and even if the data fields are used consistently across the developers.

% In this paper, we investigate the status of a bug, a data field present in most bug tracking systems that reports the progress of the bug solving activity. In particular, we focus on the activity of verification of a fix. We are interested in answering the following exploratory research questions:

% For research question xx, qualitative of a small sample. 
% For research question xx, all bugs were considered.

@@

Bug tracking systems allow users and developers of a software project to manage a list of bugs for the project. Usually, users and developers can report bugs, along with information such as steps to reproduce the bug, its severity and the operating system used. Developers choose bugs to fix and can report on the progress of the bug fixing activities, ask for clarification, discuss causes for the bug etc.

In this research, we focus on Bugzilla, an open source bug tracking system used by notable software projects such as Eclipse, Mozilla, Linux Kernel, Open Office, Apache, and companies such as NASA and Facebook\footnote{Complete list available at http://www.bugzilla.org/installation-list/.}. The general concepts from Bugzilla should apply to most other bug tracking systems.

@@

In the simplest cases, a bug is created and receive the status \texttt{UNCONFIRMED} (if it was created by a regular user) or \texttt{NEW} (if it was created by a developer). Next, it is ASSIGNED to a developer, and then it is RESOLVED, possibly by fixing it with a patch on the source code. The solution is then VERIFIED by someone in the quality assurance team. If it passes the quality requirements, it is CLOSED when the next release of the software is released. If it does not pass the quality requirements or if the solution only partially fixes the problem, the bug is REOPENed.

@@

In order to understand the bug fixing process of a software team, however, it should be possible to know what activities are performed upon change of status. For example, what steps are made before changing the status of a bug to RESOLVED (FIXED)? Is the patch applied to the version control system? Is it presented as diff file attached to the bug report? Is there a build available to end users containing the patch (e.g., a nightly build)? The documentation for Bugzilla does not prescribe specific activities for marking a bug as RESOLVED, relegating instead this decision to each software project.

Similarly, what do developers do before marking a bug as VERIFIED? Bugzilla documentation states that, when a bug is VERIFIED, it means that ``QA [quality assurance team] has looked at the bug and the resolution and agrees that the appropriate resolution has been taken''\footnote{\url{https://landfill.bugzilla.org/bugzilla-3.6-branch/page.cgi?id=fields.html}}. Again, the definition is broad, although it assumes the existence of a QA team.

@@

While mining bug repositories, one cannot assume that the VERIFIED status comprises all forms of software verification. Also, one cannot rely on the information provided by the bug tracking system or the developer documentation for a project, since it can be too generic to be useful. Such information can only be assessed by taking a close look on actual bug reports.

@@

	In order to perform the desired analysis, we needed access to all data recorded by Bugzilla for a specific project. We have found such data for two projects---Eclipse and NetBeans---from the domain of integrated development environments. The data were made available as part of the Mining Software Repositories 2011 Challenge\footnote{\url{http://www.msrconf.org/msr-challenge.html}} in the form of MySQL database dumps. The files contain all tables from the respective databases, except for the tables containing information about developers.

@@

	NetBeans\footnote{\url{http://netbeans.org/about/history.html}} started  as a student project in 1996. It was bought by Sun Microsystems in October, 1999, and it was open sourced in March, 2000.

@@

	There are 316,911 bug reports for 155 subprojects in Eclipse, and 185,578 bug reports for 39 subprojects in NetBeans.

@@

	\TODO this sampling was made just for the preliminary, qualitative analysis. Maybe we should describe preliminary analysis as an unstructured inquiry (instead of detailing the methods for it).

@@

\begin{section}{Results and Discussion}

\begin{subsection}{Preliminary Analysis}
	Some insights were gained in this step, partially answering the who, when, and how questions:

	\begin{itemize}
		\item [how] In some projects, the status VERIFIED does not correspond to software verification. In Eclipse/EMF, for example, it means that the fix is available in a build that is promoted on the project website\footnote{\url{http://wiki.eclipse.org/Modeling_PMC_Meeting,_2007-10-16}}.
		\item [(who)] In the NetBeans subprojects that were analyzed, it was common that the reporter also verified the fix. In some bugs, this was necessary since the developer was not able to reproduce the bug.
		\item [(who, when)] Often, the fixer verifies the bug just after marking it as fixed.
		\item [(when)] Sometimes, several old bugs are marked as VERIFIED at the same time.
		\item [(how?)] In Eclipse/Platform, it is common that the fixer requests someone else to verify the fix (``Please verify, Tom'').
		\item [(how)] In Eclipse/EMF, bug #249436, one developer rewrites the patch submitted by other developer.
		\item [(how)] In Eclipse/EMF, bug #269789, a test case is used to assess the proposed fix.
	\end{itemize}
\end{subsection}

@@

	In NetBeans it is possible to infer the existence of a QA team, composed by approximately 20\% of the developers, that performs about 90\% of all verifications.

	In the Eclipse subprojects, there is no evidence of a dedicated QA team. In Eclipse/EMF no developer focus specifically on verification (it should be remembered that verification in EMF has a peculiar meaning). In Eclipse/Platform, only 2\% of the developers focus on verification tasks, however, they contribute only to 1\% of the verifications.

@@

\begin{section}{Conclusion}
  The VERIFIED status is used in a variety of distinct ways, not always consistent with the traditional definition of software verification ...

	With the knowledge obtained from this exploratory research, we can improve and extend our previous work on the impact of independent verification on software quality. Examples of refined questions:

	\begin{itemize}
		\item Does independent verification improve software quality when it is performed by developers from the QA team?
		\item Does independent verification improve software quality when it is performed as an separate phase of the software release cycle?
		\item Does independent verification improve software quality when a particular verification technique (e.g., inspections, unit testing) is used?
	\end{itemize}
